Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-12-14 docs/ms_sageAbundance.Rmd
moves discussion section around

We required future projections of climate for our study area to conduct two sets of simulations: (1) comparisons of equilibrium sagebrush cover under observed climate to forecasted equilibrium cover under future climate scenarios and (2) dynamic forecasts from the present into the future.
We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.

For dynamic forecasts we used yearly GCM projections from 2012 to 2098 to make temporally-explicit forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
We aggregated daily GCM output for each GCM and RCP scenario into the yearly climate covariates used to fit our model.
These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.

@@

### Computing
####Fitting the model
We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days.
We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all paramters) [@Gelman1992; @Gelman2009].

####Simulating the process
We performed four sets of simulations to (1) compare observed and simulated equilibrium cover, (2) compare observed and simulated year- and location-specific cover, (3) forecast future equilibrium population states under projected climate change, and (4) make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098. 
Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a perturbed climate time series (for 3).
We ran equilibrium simulations (1 and 3) for 2,000 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
To calculate average future equilibrium sagebrush cover, we ran simulation (3) for each GCM and RCP scenario separately, and then averaged the results over GCMs.
For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
The random year effects were not used in any simulation.

Results
-------
Averaging across all GCMs, precipitation and temperature in our study area are projected to increase; the magnitude of increase depends on the RCP scenario (Table 1). 
Trajectories of our climate covariates from GCM projections show similar trends (Fig. 2). 
GCM projections are variable for our study area, highlighting the importance of making forecasts based on ensemble projections. [I AM TEMPTED TO CUT THIS LAST LINE]

@@

When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels. (Fig. 5).
For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig. 6).
Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).

@@

### Sagebrush response to climate and climate change
The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 3), which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013]. 
The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 3).
Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.

Our model projects only small changes in sagebrush cover in this area with future climate change. This result reflects the fact that only precipitation in the previous year emerged as an important covariate (Fig. 3) and projections of future precipitation show only modest increases (Fig. 2). Under all RCP scenarios, our model projects modest increases in sagebrush cover (Figs. 5,6).
The forecasted increases are primarily driven by a positive response to precipitation (Fig. 3), but weak effects of temperature also influenced results.
A previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline, on average, in the future [@Homer2015].
The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
First, Homer et al. (2015) used downscaled IPCC projections that allowed for spatial variation in future climate, unlike our analysis here.
Therefore, Homer et al. (2015) did predict some pixel-level increases in sagebrush cover since each pixel was subject to its own regression between climate and cover.
Second, like our study, Homer et al. (2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
Thus, the decline in precipitation led to a decline in sagebrush cover.
Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig. 6).
Lastly, Homer et al.'s (2015) analysis relied on pixel-level regressions of cover through time against climate covariates.
Our current approach is fundamentally different in that we specifically model the impact of weather on year-to-year changes in sagebrush cover.
Thus, our model captures the essential interaction point between populations and the interannual variation in weather that drives population dynamics.

>>>>

@@

### Fitting the model
We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days.
We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all paramters) [@Gelman1992; @Gelman2009].

### Simulating the process
We performed four sets of simulations to (1) compare observed and simulated equilibrium cover, (2) compare observed and simulated year- and location-specific cover, (3) forecast future equilibrium population states under projected climate change, and (4) make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098. 
Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a perturbed climate time series (for 3).
We ran equilibrium simulations (1 and 3) for 2,000 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
To calculate average future equilibrium sagebrush cover, we ran simulation (3) for each GCM and RCP scenario separately, and then averaged the results over GCMs.
For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
The random year effects were not used in any simulation.

We required future projections of climate for our study area to conduct the equilibrium and temporally-explicit forecasts described above.
We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.

For the temporally-explicit forecasts we used yearly GCM projections from 2012 to 2098 to make simulate the process starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
We aggregated daily GCM output for each GCM and RCP scenario into the yearly climate covariates used to fit our model.
These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.

Results
-------
Averaging across all GCMs, precipitation and temperature in our study area are projected to increase; the magnitude of increase depends on the RCP scenario (Table 1). 
Trajectories of our climate covariates from GCM projections show similar trends (Fig. 2).

@@

When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5).
For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig. 6).
Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).

@@

### Sagebrush response to climate and climate change
The amount of precipitation in the previous year (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change, followed by spring temperature (Fig. 3).
These results are concordant with previous research.
Wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
However, the mean estimates for the climate effects are relatively small on the standardized scale (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research on the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate or that our model is poorly specified.
More likely is that aggregated estimates of plant abundance, like percent cover, mask interannual variability at the level of the individual plant.
Such averaging inevitably reduces interannual variability, which also makes detecting drivers of internanual variability more difficult.
Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology.

Under projected climate, we forecast small positive changes in sagebrush cover in this area for all RCP scenarios (Figs. 5,6).
Our forecast is a direct result of the interaction between the estimated effect size for each cliamte covariate and the amount of change in those covariates projected in the future.
First, precipitation two years before a given year-to-year transition was the most important effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2).
Second, mean spring temperature had a positive, but small, effect on sagebrush cover changes,

This result reflects the fact that only precipitation in the previous year emerged as an important covariate (Fig. 3) and projections of future precipitation show only modest increases (Table 1, Fig. 2).
Likewise, 
While we forecast an increase in sagebrush cover at our study area, SDM studies tend to project declines in sagebrush abundance in the future due to temperature increases [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
Why do our results differ?
High elevation site...winter water good...

A previous analysis of a different subset of the remote sensing data set we used also made projections of sagebrush decline, on average, in the future [@Homer2015].
The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
First, Homer et al. (2015) used downscaled IPCC projections that allowed for spatial variation in future climate, unlike our analysis here.
Therefore, Homer et al. (2015) did predict some pixel-level increases in sagebrush cover since each pixel was subject to its own regression between climate and cover.
Second, like our study, Homer et al. (2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
Thus, the decline in precipitation led to a decline in sagebrush cover.
Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig. 6).
Lastly, Homer et al.'s (2015) analysis relied on pixel-level regressions of cover through time against climate covariates.
Our current approach is fundamentally different in that we specifically model the impact of weather on year-to-year changes in sagebrush cover.
Thus, our model captures the essential interaction point between populations and the interannual variation in weather that drives population dynamics.
